Effects of Reynolds Number on the Performance and Near-Wake Dynamics of a Vertical-Axis Cross-Flow Turbine

### 2015-10-10 paper.tex
The last commit comment should have been: Introduction - edits for first paragraph
Add line numbers and remove list of todos; resolves #20
All edits by MW from last 2 weeks. need to address To-Dos.

% Add packages and commands to include here
% The amsmath, amsthm, amssymb, hyperref, caption, float and color packages are
% loaded by the MDPI class.
\usepackage{graphicx}
%\usepackage{subfigure,psfig}
\usepackage[draft]{todonotes}
\usepackage{lineno}

@@

% Contact information of the corresponding author (Add [2] after \corres if
% there are more than one corresponding author.)
\corres{pxL3@unh.edu}

% Abstract (Less than 200 words)
\abstract{The acceptable Reynolds number mismatch for scaled physical model
testing a cross-flow turbine is investigated experimentally, both with respect
to prediction of full-scale performance, and wake characteristics, which will be
drivers of overall performance and interaction between multiple turbines.}

% Keywords: add 3 to 10 keywords
\keyword{keyword; keyword; keyword}

@@

\begin{document}
%\linenumbers
\listoftodos

\section{Introduction}

Scaled physical models are often used in science and engineering to approximate
real-world systems. However, the principle of dynamic similarity allows for
geometrically scaled systems to be dynamically identical in scale if certain
nondimensional physical parameters are matched. In the case of fluid systems,
the most important parameter is often the Reynolds number, which quantifies the
importance of viscosity on the flow physics. As such, hereafter we will use the
word ``scale'' to refer to this dynamical scale rather than the geometric scale.

With regards to wind and marine hydrokinetic (MHK) turbines, scaled physical
models are used to validate predictive numerical models, predict full-scale
performance of individual turbines, and design or investigate arrays of devices.
Despite being significantly less expensive, there is still some concern at which
scale an experiment must be performed to be relevant to full-scale application.

For numerical models, one might obtain favorable predictions for scaled systems
due to modified physics. If the numerical model is then considered
``validated,'' there is a risk that its application at full-scale may produce
incorrect predictions. For scaled physical prototypes, it is of interest how
small---since size, or scale generally correlates inversely with cost---the
prototype can be while remaining a reliable predictor of full-scale behavior. At
the very least, it is necessary to test at a scale where results can be
extrapolated, i.e. changes are occurring quasi-linearly without dramatic changes
in the underlying physical principles governing the behavior of interest, e.g.
power production in the case of a wind or MHK turbine.

Blackwell et al. investigated the effects of Reynolds number on the performance
of a 2 m diameter Darrieus vertical-axis cross-flow wind turbine with NACA 0012
blades \cite{Blackwell1976}. By varying wind tunnel speed, the turbine was made
to operate at approximately constant blade chord Reynolds number $Re_c$ ranging
from $1 \times 10^5$ to $3 \times 10^5$. In this regime the turbine power
coefficient $C_P$ was shown to be sensitive to $Re_c$, with sensitivity
diminishing at the higher Reynolds numbers, especially for turbines of lower
solidity ($Nc/R$, where $N$ is the number of blades and $R$ is the turbine's
maximum radius). More recently, Polagye and Cavagnaro observed high Reynolds
number sensitivity for a high solidity helical cross-flow turbine
\cite{Polagye2013b}, and Bravo et al. observed power coefficient of a
straight-bladed VAWT to become Reynolds number independent at $Re_c \approx 4
\times 10^5$.

When designing or studying arrays, it is common to use very small
(geometrically) scaled devices \cite{Chamorro2011, Chamorro2011b}. It is
therefore important to realize the limitations of evaluating both the power
output of turbines and the wake recovery when Reynolds number is very far from
full scale. Sometimes models used are not turbines, but wake generating objects,
e.g., porous disks, that are meant to replicate the wakes of real turbines
\cite{Goldenberg1983}. In this case, it is of interest to determine at what
scale one might be able to realistically study wake flows in an array, and also
to evaluate the effectiveness of a wake generator. In other words, a wake
generator may do a fine job simulating a scaled turbine, but how well can it
simulate a full-scale device?

Previously, is has been observed experimentally that a cross-flow turbine wake's
unique vortical mean flow field is responsible for accelerated wake recovery
when compared with conventional axial-flow propeller-type turbines
\cite{Bachant2015-JoT}. In this study we seek to replicate the same energy
balance considerations at multiple Reynolds numbers, to examine the implications
on how small scale, i.e. low Reynolds number experiments may be used to study
flows in turbine arrays.

The Reynolds number effects on the maximum performance and 2-D near-wake
profiles of the University of New Hampshire's Reference Vertical Axis Turbine
(UNH-RVAT) were presented \cite{Bachant2014}. In this study we set out to
analyze in greater depth the effects of scale, i.e. Reynolds number, on the
performance of the turbine, and its more detailed near-wake characteristics. We
were looking to find a threshold scale beyond which a physical model study could
reliably predict full-scale performance. We were also looking to investigate the
sensitivity as small scales, and what the implications might be for validating
the prediction capabilities of numerical models.

\todo[inline]{Talk about interdependence of Froude number.}

@@

In this study we are focused on a turbine constructed from foils---a lift-driven
device in contrast to drag-driven. A logical starting point for understanding
the effects of $Re$ on overall turbine loading is to look at simplified cases
of individual foils, with respect to their static and dynamic loading.

@@

Blade performance is generally characterized by the lift-to-drag ratio $l/d$,
which is a function of the foil's angle of attack $\alpha$. The angles of attack
for cross-flow turbine blades are a function of turbine tip speed ratio,
induction (slowing/turning of the free stream flow prior to reaching the
turbine), and turbine rotation or azimuthal angle $\theta$. Looking at static
airfoil data, we see that the static stall angle---the angle just beyond that at
which $l/d$ is maximum---increases with blade chord Reynolds number $Re_c$. This
in general is caused by the blade boundary becoming more turbulent as a fraction
of the chord length, diminishes the negative effects of a laminar separation
bubble occurring near the leading edge. For smooth airfoils, this transition can
cause a dramatic increase in foil performance, a shown in
Figure~\ref{fig:McMasters} at a blade chord Reynolds number on the order of
$10^5$ \cite{McMasters1980}. A review of low Reynolds number airfoil behavior is
presented in \cite{Lissaman1983}. Note that there is a distinct lack of highly
reliable data for airfoils in this transitional regime and below. An evaluation
of the various databases relevant to cross-flow turbines is presented in
\cite{Bedon2014}.

@@

Static foil performance does not tell the whole story for a cross-flow turbine.
The azimuthal, and therefore temporal variation of $\alpha$ in a cross-flow
turbine implies dynamic loading. Furthermore, the blades commonly undergo
dynamic stall for tip speed ratios at and below those of maximum rotor
torque\cite{Para2002}. We then must understand the effects of Reynolds number on
the dynamic stall process.

@@

Bousman showed that dynamic stall was insensitive to Reynolds number for $Re=1.0
\times 10^5$--$2.5 \times 10^5$ \cite{Bousman2000-evaluation}, Though Singleton
and Yeager Jr. state that Reynolds number effects on dynamics stall remains an
unsolved question \cite{Singleton2000}.

@@

\todo[inline]{What do studies in the literature tell us about $Re$ effects
on shear flows, wakes, and vortex flows?}

Chamorro et al. showed that turbulence statistics in an AFT wake became
$Re$-independent at $Re_D \approx 1 \times 10^5$, with mean velocity slightly
earlier at $Re_D \approx 5 \times 10^4$ \cite{Chamorro2012}.

\section{Experimental Setup}

Experiments were performed in the University of New Hampshire's tow/wave tank
turbine test bed. The turbine model is dubbed the UNH-RVAT, for Reference
Vertical Axis Turbine, which is designed to be a generic case for numerical
model testing, similar to the Sandia National Labs/US Department of Energy RM2
River Turbine, but with higher solidity, or blade chord to radius ratio. A CAD
model of the turbine, depicted in \ref{fig:turbine}, is available from
\cite{Bachant2014-RVAT-CAD}.

\begin{figure}[ht!]
\centering
\includegraphics[width=0.4\textwidth]{figures/turbine}
\caption{Turbine model drawing. Note that the upper and lower mounting
flanges have been excluded as these were included in the tare drag
measurements.} 
\label{fig:turbine}
\end{figure}

The turbine was mounted in a frame constructed from NACA 0020 sections. The
turbine shaft ran up through the water surface, coupled to a permanent magnet
servo motor with a 20:1 gearbox. This servo was controlled by the tow tank's
main motion controller for high synchronization with the carriage motion. A
rotary torque transducer was installed inline between the servo and the turbine,
and the servo was also mounted on a slewing ring bearing, which allowed a
redundant measurement of torque via an arm and load cell used to counteract the
turbine moment. The frame was mounted to the carriage via linear guides, such
that the total streamwise drag force was transferred to a pair of S-beam load
cells, providing the rotor drag measurements, after a measured tare drag was
subtracted in post processing. Similarly, a tare torque was measured by rotating
the turbine shaft in air. Turbine angular and tow carriage linear position were
measured using quadrature encoder signals, with $10^5$ counts-per-rev for the
turbine and 10 $\mathrm{\mu m}$ resolution for the carriage position. These
signals, along with the torque and drag signals, were sampled at 2 kHz.

\begin{figure}[ht!]
\centering
\includegraphics[width=0.75\textwidth]{figures/exp_setup_drawing}
\caption{Drawing of the experimental setup.}
\label{fig:exp-setup}
\end{figure}

Wake velocity was measured using a Nortek Vectrino+ acoustic Doppler velocimeter
(ADV), sampling at 200 Hz. The probe was mounted on an automated positioning
system, also controlled by the tow tank's main motion controller. The ADV and
DAQ systems' sampling times were synchronized by triggering the start of data
acquisition via a pulse sent from the motion controller.

@@

Approximately 1500 tows were performed, each one used for a single data point on
either a performance curve or wake map. A performance curve consisted of 31
tows, where during each tow the mean turbine tip speed ratio was held constant,
ranging from 0.1--3.1 in 0.1 increments. Full performance curve data were
acquired for tow speeds from 0.4 to 1.2 m/s in 0.2 m/s increments, for which
turbine diameter and approximate blade chord Reynolds number are presented in
Table~\ref{tab:Re}. Performance was also measured for $\lambda=1.9$ at tow
speeds $[0.3, 0.5, 0.7, 0.9, 1.1, 1.3]$ m/s for two tows each.

@@

A wake map was generated by positioning the Vectrino probe at 270 different
locations, varied in the cross-stream and vertical directions at one turbine
diameter downstream ($x/D=1$). These locations have vertical coordinates from
the turbine centerline up to $z/H=0.625$, ranging in the cross-stream direction
$y/R = \pm 3$. These locations are shown in Figure~\ref{fig:wake-locations}.

@@

@@

From each set of tows, a standard time interval was set, which allowed the
turbine performance and wake to reach a quasi-periodic state. Each run was
analyzed to compute statistics over this interval, truncating the end slightly
to achieve and integer number of blade passages. Turbine shaft angular velocity
and tow carriage speed were calculated using a second order central differencing
scheme on the respective position measurements. Power and drag coefficients are
calculated as instantaneous quantities using the carriage speed as the free
stream velocity.

Wake velocity data were filtered for spurious ``spikes'' by removing datapoints
8 standard deviations or 0.9 m/s from the mean. The experimental data and code
for processing and visualization are available from
\cite{Bachant2015-RVAT-Re-dep-data}.

@@

Full power and rotor drag coefficient curves for various Reynolds numbers are
plotted in Figure~\ref{fig:cp-curves} and Figure~\ref{fig:cd-curves},
respectively. There is essentially no change in the shape of the drag
coefficient curves---merely an upward shift in $C_D$ with increasing $Re$.

In general, maximum $C_P$ increases with Reynolds number, which is due to an
increase in the foil lift-to-drag ratio. The power coefficient curves also show
a downward shift in the optimal tip speed ratio with increasing Reynolds number.
This is caused by the stall delay from a more turbulent boundary later on the
blade suction side.

\begin{figure}[ht]
\centering
\includegraphics[width=0.65\textwidth]{figures/cp_curves}
\caption{Mean power coefficient curves plotted for
multiple Reynolds numbers.}
\label{fig:cp-curves}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=0.65\textwidth]{figures/cd_curves}
\caption{Mean rotor drag coefficient curves plotted for
multiple Reynolds numbers.}
\label{fig:cd-curves}
\end{figure}

Mean power and drag coefficients at $\lambda=1.9$ are plotted for all Reynolds
numbers in Figure~\ref{fig:perf-Re-dep}. There is a drastic improvement in $C_P$
with increasing Reynolds number for the lower values. Power coefficient then
becomes  essentially $Re$-independent at $Re_D = 0.8 \times 10^6$, which
corresponds to an approximate average blade chord Reynolds number $Re_{c,
\mathrm{ave}} = 2.1 \times 10^5$. This threshold is consistent with the behavior
of the blade boundary layer transitioning from laminar to turbulent, thereby
promoting reattachment of the laminar separation bubble.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.95\textwidth]{figures/perf_re_dep}
\caption{Mean power (left) and drag (right) coefficients at $\lambda=1.9$
plotted versus Reynolds number.} 
\label{fig:perf-Re-dep}
\end{figure}

@@

To help understand---and possibly predict---the $Re$-sensitivity on turbine
performance, a series of static foil coefficient datasets were computed with the
XFOIL viscous panel method code \cite{Drela1989}, implemented as part of the
open source turbine design software QBlade. Simulations were run for an angle of
attack range of 0--40 degrees, in increments of 0.5 degrees. Solver parameters
used a 100 panels, a fixed speed, zero Mach number, \texttt{NCrit=9} (default
$\mathrm{e}^n$ transition criteria parameter for an average wind tunnel), and no
forced boundary layer transition. Characteristics for computed for the
approximate average blade chord Reynolds numbers encountered in the tow tank
experiments.

@@

\begin{figure}[ht!]
\centering
\includegraphics[width=0.9\textwidth]{figures/all_foils_re_dep}
\caption{Foil section characteristics computed by XFOIL for various $Re_c$.}
\label{fig:foil-Re-dep}
\end{figure}

In conjunction with the cross-flow turbine blade kinematics, the foil
coefficients were used to approximate turbine performance by calculating the
peak torque coefficient on the upstream half of the blade path. The turbine
torque coefficient $C_T$ can be related to the blade section chordwise force
coefficient $C_c$ by
\begin{equation}
C_T = \frac{C_c c}{2R} \frac{|U_\mathrm{rel}|^2}{U_\infty^2},
\end{equation}
where the blade section chordwise force (for zero preset pitch)
\begin{equation}
C_c = C_l \sin \alpha - C_d \cos \alpha.
\end{equation}

The relative blade velocity was calculated by adding the free stream to the
opposite of the blade tangential velocity. Note that this neglects any
induction, or slowing of the free stream by the turbine forces, which would be
present in a momentum/streamtube model. Since we were not looking to predict
absolute performance but rather relative changes with $Re$, this method was
deemed acceptable as it is extremely simple and fast to compute.

Values for the blade angle of attack, relative velocity, and torque coefficient
are plotted in Figure~\ref{fig:blade-kinematics}. The effects of static stall
are clearly present in the torque coefficient plot, and by the time the angle of
attack has decreased below that of stall, the relative velocity is so low that
there is not much contribution to the torque coefficient.

\begin{figure}[ht!]
\centering
\includegraphics[width=0.9\textwidth]{figures/foil_kinematics_ct}
\caption{Geometric angle of attack, relative velocity and torque coefficient
calculated with a NACA 0020 foil.}
\label{fig:blade-kinematics}
\end{figure}

Results for the peak torque coefficient for each foil are plotted versus $Re_c$
in Figure~\ref{fig:foils-C_T-Re-dep}. It is interesting that the convergence of
$C_{T_\mathrm{max}}$ with increasing Reynolds number is more dramatic than any of the
common foil characteristics plotted in Figure~\ref{fig:foil-Re-dep}, meaning
that the cross-flow turbine's unique kinematics must be taken into account when
attempting to predict the effect of transitional Reynolds numbers on turbine
performance.

From the peak torque coefficient metric, the Reynolds number independence is
achieved at lower values, and more dramatically. We see that the trend of the
NACA 0020 curve matches almost perfectly at low $Re$, but the trend to continue
increasing linearly is not matched by the experimental data, which looks to
behave more like the cambered foil results. Though this method is very
approximate, it nevertheless appears to be a reasonable way to predict the
transitional Reynolds number regime for cross-flow turbine performance using
only 2-D static airfoil characteristics

>>>>

% Add packages and commands to include here
% The amsmath, amsthm, amssymb, hyperref, caption, float and color packages are
% loaded by the MDPI class.
\usepackage{graphicx}
%\usepackage{subfigure,psfig}
\usepackage[draft]{todonotes}

@@

% Contact information of the corresponding author (Add [2] after \corres if
% there are more than one corresponding author.)
\corres[2]{pxL3@unh.edu, martin.wosnik@unh.edu}

% Abstract (Less than 200 words)
\abstract{The performance and wake characteristics of cross-flow (vertical 
axis) turbines depend on many parameters: solidity, blade profile, blade pitch, 
number of blades, strut drag, operational parameters such as tip speed ratio 
--- and Reynolds number. If numerical models are validated with physical model 
data that was obtained at insufficiently high Reynolds numbers, it cannot be 
determined whether problems with model predictions are caused by Reynolds 
number effects, other turbine parameters or issues with the numerical model.
%It is uncertain whether numerical models validated with physical model data 
%obtained at low Reynolds number should be considered validated at all, since 
%the scale at which the model will be applied is orders of magnitude larger. 
Experiments were performed with a large laboratory-scale cross flow turbine of 
to investigate Reynolds number dependence of performance and wake transport 
characteristics, 
and to determine what an acceptable Reynolds number mismatch is for scaled 
physical model tests of cross-flow turbines. It was demonstrated that the 
performance of the cross-flow turbine becomes essentially $Re$-independent at a 
Reynolds number based on turbine 
diameter of $Re_D \approx 10^6$, or an approximate Reynolds number based on 
blade chord of $Re_c \approx 2 \times 10^5$. 
A simple model that calculates peak torque coefficient from 
static foil data and cross-flow turbine kinematics was shown to be a reasonable 
estimator for predicting Reynolds number dependence of an actual cross flow 
turbine operating under dynamic conditions.}

% Keywords: add 3 to 10 keywords
\keyword{Reynolds number; cross-flow turbine; turbine performance; marine 
hydrokinetic energy; wind energy}

@@

\begin{document}
\todo[inline]{Rewrite abstract so it is more of a summary of what was done - 
tone down the background information a bit. (Note: some background information 
was moved to Introduction)} 
% %
\todo[inline]{Include a List of Symbols?} 
% %

% %
\listoftodos
% %

\section{Introduction}
Scaled physical models are often used in science and engineering to approximate 
real-world systems. The principle of dynamic similarity allows for
geometrically scaled systems to be dynamically identical in scale if certain
nondimensional physical parameters are matched. In the case of fluid systems,
the most important parameter is often the Reynolds number, which quantifies the
importance of viscosity on the flow. As such, hereafter we will use the word
``scale'' to refer to this dynamical scale rather than the geometric scale.

With regards to wind and marine hydrokinetic (MHK) turbines, scaled physical
models are used to validate predictive numerical models, predict full-scale
performance of individual turbines, and design or investigate arrays of devices.
Scaled models have the benefit of being significantly less expensive, however, 
there still is concern at which
scale an experiment must be performed to be relevant to full-scale application. 
This is particularly important for cross-flow turbines, where Reynolds number 
effects for experiments conducted at low $Re$ are often difficult to 
distinguish from the effects on performance of several other turbine 
parameters.

The performance and wake characteristics of cross-flow (vertical 
axis) turbines depend on turbine solidity, blade profile (lift/drag, dynamic 
stall at reduced frequency of turbine rotation, symmetry, thickness, camber), 
blade pitch, number of blades, strut drag and operational parameters such as 
tip speed ratio --- {\em and} on Reynolds number. 
Note that an average blade chord Reynolds number, $Re_{c,\mathrm{avg}} \approx 
\lambda U_\infty c/ \nu$, can be expressed in terms of tip speed ratio $\lambda 
\equiv \omega R/ U_\infty$; the value of $\lambda$ at which a turbine reaches 
peak performance is again a function of turbine solidity $Nc/\pi D$.

If numerical models are validated with physical model data that
was obtained at insufficiently high Reynolds numbers, it cannot be determined
whether problems with model predictions are caused by Reynolds number effects, 
other turbine parameters or issues with the numerical model.
One might also obtain different predictions of scaled systems
due to modified physics at lower Reynolds numbers, e.g., laminar boundary 
layers and transitional flow regimes. If the numerical model is then considered
``validated,'' there is a risk that its application at full-scale may produce
incorrect predictions. It is uncertain whether numerical models validated with 
physical model data obtained at low Reynolds number should be considered 
validated at all, since the scale at which the model will be applied is orders 
of magnitude larger.

This can be overcome by showing that a scaled physical model test has become 
Reynolds number independent, and therefore validation efforts should be 
relevant at full-scale.  For scaled physical models, it is of interest how
small---since size, or scale generally correlates inversely with cost---the
model can be while remaining a reliable predictor of full-scale behavior. At
the very least, it is necessary to test at a scale where results can be
extrapolated, i.e., changes are occurring quasi-linearly without dramatic 
changes in the underlying physical principles governing the behavior of 
interest, e.g., power production in the case of a wind or MHK turbine.

Blackwell et al. investigated the effects of Reynolds number on the 
performance
of a 2 m diameter Darrieus vertical-axis cross-flow wind turbine with NACA 0012
blades \cite{Blackwell1976}. By varying wind tunnel speed, the turbine was made
to operate at approximately constant blade chord Reynolds number $Re_c$ ranging
from $1 \times 10^5$ to $3 \times 10^5$. In this regime the turbine power
coefficient $C_P$ was shown to be sensitive to $Re_c$, with sensitivity
diminishing at the higher Reynolds numbers, especially for turbines of lower
solidity ($Nc/R$, where $N$ is the number of blades and $R$ is the turbine's
maximum radius). More recently, Bravo et al. observed power coefficient of a
straight-bladed VAWT to become Reynolds number independent at $Re_c \approx 4
\times 10^5$.
\todo[inline]{Add reference for Bravo et al.} 
\todo[inline]{Why did you take Polagye and Cavagnaro citation out - not 
relevant?} 
\todo[inline]{bring back the examples of studies at small scale? (too small)}

When designing or studying turbine arrays, it is common to use very small
(geometrically) scaled devices. 
% %
\todo[inline]{Get a reference for this, and a sample geometric scale.} 
% %
Sometimes these are not even models of turbines, but
wake generating objects, e.g., porous disks, that are meant to replicate the
wakes of real turbines. In this case, it is of interest to determine at what
scale one might be able to realistically study wake flows in an array, and also
to evaluate the effectiveness of a wake generator. In other words, a scale 
model wake generator may do a fine job simulating a model turbine of the same 
scale, but how well do either of them simulate a real turbine?
\todo[inline]{examples of porous disk papers, e.g. Aubrun et al?}

Previously, Bachant and Wosnik \cite{Bachant2015-JoT} showed that a cross-flow
turbine wake's unique vertical mean flow field is responsible for accelerated
wake recovery when compared with conventional axial-flow propeller-type
turbines. In this study we seek to replicate the same energy balance
considerations at multiple Reynolds numbers, to examine the implications on how
small scale, i.e. low Reynolds number experiments may be used to study flows in
turbine arrays.

Bachant and Wosnik \cite{Bachant2014} observed that maximum performance and 2-D 
near-wake profiles are affected by Reynolds number. In the present study we set 
out to analyze in greater depth the effects of scale, i.e. Reynolds number, on 
the performance of the turbine and its detailed near-wake 
characteristics. 
The goal of this study was to find a threshold scale beyond which a physical 
model study could reliably predict full-scale performance. 
%We were also looking to investigate the sensitivity as small scales, and what 
%%%the implications might be for validating
%the prediction capabilities of numerical models.
Experiments were performed that investigated Reynolds number dependence for 
scaled physical model tests of a cross-flow turbine with respect
to prediction of full-scale performance, and wake characteristics, which will be
drivers of overall performance and interaction between multiple 
turbines.

\todo[inline]{Talk about interdependence of Froude number. whatis teh Froude 
number at 1 m/s?}

@@

In this study we are using a simple cross-flow turbine constructed from 
hydrofoils, i.e., the 
turbine is a lift-driven device in contrast to drag-driven. A logical starting 
point for understanding the effects of $Re$ on overall turbine loading is to 
first consider simplified cases of individual foils, and their static and 
dynamic loading.

@@

Hydro- or airfoil section performance is generally characterized by the 
lift-to-drag 
force ratio $L/D$  (or lift vs drag coefficient, $c_L/c_D$ ),
which is primarily a function of the foil's angle of attack $\alpha$. The 
angles of attack
for cross-flow turbine blades are a function of turbine tip speed ratio,
induction (slowing/turning of the free stream flow prior to reaching the
turbine), and turbine rotation or azimuthal angle $\theta$. Looking at static
airfoil data, we see that the static stall angle---the angle just beyond that at
which $L/D$ is maximum---increases with blade chord Reynolds number $Re_c$ 
(reference?). This
is caused by the boundary layer on the foil becoming more turbulent and more 
energetic, which enables it to advance further downstream against the adverse 
pressure gradient on the suction side and delay separation to higher angels of 
attack, as well as diminish the negative effects of a laminar separation
bubble occurring near the leading edge (reference?). For smooth airfoils, this 
transition can
cause a dramatic increase in foil performance, as seen in
Figure~\ref{fig:McMasters} at a blade chord Reynolds number on the order of
$10^5$ \cite{McMasters1980}. A review of low Reynolds number airfoil behavior is
presented in \cite{Lissaman1983} and corroborates these trends.

\todo[inline]{Laminar separation bubble - is it always there? alse need 
reference}

@@

Static foil performance does not tell the whole story for a cross-flow turbine.
The azimuthal, and therefore temporal variation of $\alpha$ in a cross-flow
turbine implies dynamic loading. Furthermore, since the range of angle of 
attack encountered by a rotating blade increases with decreasing tip speed 
ratio, the blades commonly undergo
dynamic stall for tip speed ratios at and below those of maximum rotor
torque\cite{Para2002}. We then must understand the effects of Reynolds number on
the dynamic stall process.

@@

Bousman showed that dynamic stall was insensitive to Reynolds number for $Re=1.0
\times 10^5$--$2.5 \times 10^5$ \cite{Bousman2000-evaluation}.

\todo[inline]{Bousman: explain in what type of system, with what type of foil 
this was done.}
\todo[inline]{you took quite a few references out of this section...?}
\todo[inline]{additional data in helicopter literature?}

@@

\todo[inline]{What do studies in the literature tell us about $Re$ effects
on shear flows, wakes, and vortex flows? Note-MW I'm looking into Re 
dependence in general wake studies.}

Chamorro et al. \cite{Chamorro2012} showed that turbulence statistics in an AFT 
wake became
$Re$-independent at $Re_D \approx 1 \times 10^5$, and that mean velocity 
profiles became $Re$-independent slightly earlier at $Re_D \approx 5 \times 
10^4$.
\todo[inline]{What was Chamorro's blade chord Re?}
\section{Experimental Setup}

Experiments were performed in the University of New Hampshire's tow/wave tank
in a turbine test bed specifically designed for cross-flow turbines. The 
turbine model used in this study was the UNH Reference Vertical Axis Turbine 
(RVAT), which was designed to be a generic test case for numerical
model validation, similar to the Sandia National Labs/US Department of Energy 
RM2
River Turbine, but with somewhat higher solidity, or blade chord to radius 
ratio.

\todo[inline]{add reference: V. Neary, M. Previsic, R. Jepsen, M. Lawson, Y. 
Yu, A. Copping, A. Fontaine, K. Hallett, D. Murray (2014). Methodology for 
Design and Economic Analysis of Marine Energy Conversion (MEC) Technologies. 
SAND2014-9040}

The UNH-RVAT turbine has three blades made from NACA 0020 profiles with 
0.140 m chord. The blades are mounted at mid-chord w.r.t. the turbine axis, and 
the turbine has a height (blade span) of 1 m and a diameter of 1 m, c.f. 
Figure~\ref{fig:turbine}. The turbine test bed allows precise control and 
measurement of the turbine shaft angle and angular velocity, and therefore tip 
speed ratio, through a permanent magnet servo motor.
The primary torque measurement was provided by an in-line torque cell, with a 
redundant measurement by a moment arm and load cell and an indirect measurement 
through the motor via its torque constant.
The turbine was mounted with in a low-drag frame constructed from NACA 0020 
foil sections.  The frame is attached to the tow carriage by four linear 
bearings, allowing the total streamwise drag force to be measured by a pair of 
S-beam load cells. Turbine rotor drag is calculated by subtracting the test bed 
tare drag.
A schematic of the turbine test bed is shown in 
Figure~\ref{fig:exp-setup}.
Additional details of the turbine and experimental setup are described in
\cite{Bachant2015-JoT}, and a CAD model of the turbine is available from
\cite{Bachant2014-RVAT-CAD}.

\begin{figure}[ht!]
\centering
\includegraphics[width=0.4\textwidth]{figures/turbine}
\caption{UNH-RVAT turbine model. Turbine blades and struts made from NACA 0020 
profiles with 0.140 m chord. Note that the upper and lower mounting
flanges have been excluded as these were included in the tare drag
measurements.} 
\label{fig:turbine}
\end{figure}

\begin{figure}[ht!]
\centering
\includegraphics[width=0.75\textwidth]{figures/exp_setup_drawing}
\caption{Experimental setup: turbine test bed installed in UNH tow tank.}
\label{fig:exp-setup}
\end{figure}

@@

Approximately 1500 tows were conducted for the study reported here, each tow 
was used for a single data 
point on either a performance curve or wake map. 
Each performance curve consisted of 31
tows, where during each tow the mean turbine tip speed ratio was held constant,
ranging from 0.1--3.1 in 0.1 increments. Full performance curve data were
acquired for tow speeds from 0.4 to 1.2 m/s in 0.2 m/s increments, for which
turbine diameter and approximate blade chord Reynolds number are presented in
Table~\ref{tab:Re}. Performance was also measured for $\lambda=1.9$ at tow
speeds $[0.3, 0.5, 0.7, 0.9, 1.1, 1.3]$ m/s for two tows each.

@@

Each wake map was generated by positioning a Nortek Vectrino+ acoustic Doppler
Velocimeter (ADV), measuring three components of velocity at 200 Hz sampling 
frequency, at 270 different locations, varied in the cross-stream and
vertical directions at one turbine diameter downstream ($x/D=1$). These
locations have vertical coordinates from the turbine centerline up to
$z/H=0.625$, ranging in the cross-stream direction $y/R = \pm 3$. These
locations are shown in Figure~\ref{fig:wake-locations}.

@@

\todo[inline]{Add a thicker dashed line at $y= -0.5m$, and maybe a dashed 1m 
circle in the top view of the turbine.}

@@

For each set of tows, a standard time interval was set, and each run was
analyzed to compute statistics over this interval, truncating the end slightly
to achieve an integer number of blade passages.

The experimental data and processing/plotting code are available from
\cite{Bachant2015-RVAT-Re-dep-data}.

@@

Full performance and turbine rotor drag )thrust) curves for various Reynolds 
numbers are plotted in Figure~\ref{fig:perf-curves}. 
In general, maximum $C_P$ increases with Reynolds number, from about 0.20 at 
$Re_D=0.4 \times 10^6$, to 0.26 at higher Reynolds numbers, which is due to an
increase in the foil lift-to-drag ratio. The power coefficient curves also show
a slight downward shift in the optimal tip speed ratio (peak performance) with 
increasing Reynolds 
number, from about $\lambda \approx$ 2.0 to 1.9.
This is caused by the stall delay from a more turbulent boundary later on the
blade suction side.
There is essentially no change in the shape of the
drag coefficient curves---merely an upward shift in $C_D$ with increasing $Re$.

\begin{figure}[ht]
\includegraphics[width=0.95\textwidth]{figures/perf_curves}
\caption{UNH-RVAT experiments: mean power (left) and rotor drag (right) 
coefficient curves plotted for multiple Reynolds numbers.}
\label{fig:perf-curves}
\end{figure}

Mean power and drag coefficients at $\lambda=1.9$ are plotted vs.\ Reynolds
number in Figure~\ref{fig:perf-Re-dep}. There is a drastic improvement in $C_P$
with increasing Reynolds number at lower values of Reynolds number. The power 
coefficient then
becomes  essentially $Re$-independent at $Re_D = 0.8 \times 10^6$, which
corresponds to an approximate average blade chord Reynolds number $Re_{c,
\mathrm{ave}} = 2.1 \times 10^5$. This threshold is consistent with the behavior
of the blade boundary layer transitioning from laminar to turbulent, thereby
promoting reattachment of the laminar separation bubble.
\todo[inline]{Is there still a laminar separation bubble near the leading edge 
at these Reynolds numbers?}

@@

\begin{figure}[ht]
\includegraphics[width=0.95\textwidth]{figures/perf_re_dep}
\caption{UNH-RVAT experiments: mean power (left) and rotor drag (right) 
coefficients at $\lambda=1.9$ plotted versus Reynolds number.} 
\label{fig:perf-Re-dep}
\end{figure}

@@

To help understand---and possibly predict---the $Re$-sensitivity on turbine
performance, a series of static foil coefficient datasets were computed with the
XFOIL viscous panel method code \cite{Drela1989}, implemented as part of the
open source turbine design software QBlade. Simulations were run with the
default settings and zero Mach number, for the range of blade chord
Reynolds numbers encountered in the experiments.

@@

\begin{figure}[ht!]
\centering
\includegraphics[width=0.95\textwidth]{figures/all_foils_re_dep}
\caption{Foil section characteristics computed by XFOIL for various $Re_c$.}
\label{fig:foil-Re-dep}
\end{figure}

In conjunction with the cross-flow turbine blade kinematics, the foil
coefficients were used to approximate turbine performance by calculating the
peak torque coefficient on the upstream half of the blade path. The turbine
torque coefficient $C_T$ can be related to the blade section chordwise force
coefficient $C_c$ by
\begin{equation}
C_T = \frac{C_c c}{2R} \frac{|U_\mathrm{rel}|^2}{U_\infty^2},
\end{equation}
where the blade section chordwise force (for zero preset pitch) is given by
\begin{equation}
C_c = C_l \sin \alpha - C_d \cos \alpha.
\end{equation}

The relative blade velocity $U_\mathrm{rel}$ was calculated by vector addition 
of the free 
stream velocity and the opposite of the blade tangential velocity. Note that 
this 
neglects any induction, or slowing of the free stream by the turbine forces, 
which would be present in a momentum/streamtube model. Since the goal of this 
approach was not to predict absolute performance but rather gain insight into 
relative changes with $Re$, 
this method was deemed acceptable as it is extremely simple and fast to 
compute.

Values for the blade angle of attack, relative velocity, and torque coefficient
are plotted in Figure~\ref{fig:blade-kinematics} for the upstream half of 
rotation of the turbine. The effects of static stall
are clearly present in the torque coefficient plot, and by the time the angle of
attack has decreased below that of stall, the relative velocity is so low that
there is not much contribution to the torque coefficient.

\begin{figure}[ht!]
\centering
\includegraphics[width=0.95\textwidth]{figures/foil_kinematics_ct}
\caption{Geometric angle of attack (a), relative velocity (b) and torque coefficient
(c) calculated with a NACA 0020 foil.}
\label{fig:blade-kinematics}
\end{figure}

Results for the normalized peak torque coefficient for each foil are plotted 
versus $Re_c$
in Figure~\ref{fig:foils-C_T-Re-dep}. It is interesting that the convergence of
$C_{T_\mathrm{max}}$ with increasing Reynolds number is more dramatic than any of the
common foil performance characteristics plotted in 
Figure~\ref{fig:foil-Re-dep}, meaning
that the cross-flow turbine's unique kinematics must be taken into account when
attempting to predict the effect of transitional Reynolds numbers on turbine
performance.
